\section*{Week 3: Inventory and IFRS; Long-Term Assets}

\subsection*{The Inventory Equation}

$BegInventory + Additions = COGS + EndInventory$ \

\subsection*{US LIFO Conformity Rule}
LIFO for tax purposes $\implies$ LIFO for financial reporting \
FIFO for tax purposes $\implies$ FIFO for financial reporting \

\subsection*{Comparability}
LIFO and FIFO firms have different account, so we need to adjust the accounting numbers to make them comparable.  Data is available to adjust the LIFO firm to FIFO (but not to adjust FIFO to LIFO)

$LIFO Reserve = EndingInv_{FIFO} - EndingInv_{LIFO}$ \\

Derivation of adjustment LIFO to FIFO:\\
(1): $EndInv_{FIFO} = BegInv_{FIFO} + Additions - COGS_{FIFO}$  \\
(2): $EndInv_{LIFO} = BegInv_{LIFO} + Additions - COGS_{LIFO}$  \\
(1)-(2):\\
$EndInv_{FIFO} -EndInv_{LIFO} = BegInv_{FIFO} - BegInv_{LIFO} - COGS_{FIFO} - - COGS_{LIFO}$ \\
$EndLIFOReserve = BegLIFOReserve+ COGS_{LIFO} - COGS_{FIFO}$ \\

$\Delta LIFOReserve = COGS_{LIFO} - COGS_{FIFO}$ \\

$ COGS_{FIFO} = COGS_{LIFO} - \Delta LIFOReserve  $


\subsection*{Ratios Involving Intentory}

$InventoryTurnover = \frac{COGS}{AverageInventory}$ \

$DaysInventory = \frac{365}{InventoryTurnover}$ \

\subsection*{FIFO/LIFO Summary}

\begin{itemize}
	\item LIFO results in a more accurate income statement; IFO results in a more accurate balance sheet.
	\item LIFO-FIFO is import to compare across firms and adjust the LIFO reserve for comparability.
\end{itemize}


\subsection*{PPE: Property, Plant and Equipment}

\begin{enumerate}
	\item Whats the acquisition cost?
	\item How much is the estimated salvage value? (value at disposal net of selling costs)
	\item What is the expected useful service life? (period of usage not physical life)
	\item What is the depreciation pattern?
\end{enumerate}
	
Straight Line Method of Depreciation : 

$ExpensePerYear = \frac{AcquisitionCost - SalvageValue}{EstimatedUsefulLife}$ \


\subsection*{Gain or Loss on Sale or Disposal of PPE}
Procedure for sale or disposal of PPE:
\begin{enumerate}
	\item Record cash or the market value of the asset received for the PPE
	\item Record disposal of the asset by removing the cost of the asset from PPE
	\item Remove the accumulated depreciation associated with the asset
	\item Calculate gain or loss as follows: \\  
	$ Cash - (Cost - AccDep) = Gain (Loss) $
\end{enumerate}


Disposal of asset G/L: $G/L = Cash-(Cost-AccDep)$ \
 $G/L = SalesPrice-(GrosPPE-AccDep)$ \
 $G/L = AccDep + (GrossPPE-SalesPrice)$ \
 $G/L = AccountingDepreciation - EconomicDepreciation$


We impair or write down the asset by increasing accumulated depreciation.

$Acquisitions = Ending PP\&E - Beginning PP\&E + PP\&E Sold$

$Beginning accumulated depreciation + Depreciation expense - depreciation on sold assets = Ending accumulated depreciation$ \\

$Gain (Loss) = Sale price - Book Value$ \\
$Book Value = Purchase price - Accumulated Depreciation$
